GEE models

### 2016-05-04 doc/manuscript.Rmd
Added Richard's edits/suggestions, sent to Tony

@@

Title: Total non-esterified fatty acids, but not individual species, predict
progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

Word count: 3543 / 4000

@@

# Abstract
250 / 250 words

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acid (NEFA) species on changes in insulin sensitivity and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

@@

Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
6 years. Our results suggest that association with beta-cell function is due to
the absolute size of the serum NEFA fraction, rather than the specific lipid
composition.

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
- [X] Replace TAG with BaseTAG in the QIC (or use both)
- [X] Fix time axis labels on LCMM plot
- Describe LCMM and PLS-DA better in plots

@@

sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')

@@

sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.')

@@

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
Epidemiological and clinical studies have reported concordant findings, showing
in particular that elevated total plasma NEFA associates with an increased risk
for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual lipid species such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple classes of molecules with diverse physiological functions, and few 
studies have compared the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in the research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
species [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the NEFA
blood lipid fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual lipid species. Therefore, our research objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and triacylglycerol) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

The primary outcome variables for this analysis were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA species as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models (M5-M8) were comparable in performance, so we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we a priori
decided to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TG is a risk factor
for diabetes and that NEFA contribute to TG production, TG may act as a mediator
between NEFA and the outcomes. To determine the role of TG in the
association between NEFA and the outcomes, TG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

@@

Several associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for the concentration compared to the
mol% data. Full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
had negative associations between
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD decrease
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuated all associations with IGI/IR and ISSI-2
(data not shown).
There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (all p>0.15). Raw values from the GEE models are shown in 
`r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee')` for fully
adjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_groups')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_loadings')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants
were classified correctly based on dysglycemia status, examining the clustering
of the NEFA composition showed poor discrimination between groups (see 
`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
total NEFA had a negative association with insulin sensitivity and a
particularly strong negative association with beta-cell function [@Johns2014a].
Experimentally, several potential mechanisms have been elucidated for the role
of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
through endoplasmic reticulum stress, formation of ceramides, and generation of
nitric oxide, as well as impairment of proinsulin production and mitochrondrial
function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
The present analysis is the first study to our knowledge to examine the 
longitudinal association of a broad spectrum of individual NEFA species on
beta-cell function in a large cohort. We found that there was a strong signal of
higher total NEFA, palmitic acid, oleic acid, and linoleic acid as
concentrations with lower beta-cell function. However, in modeling these fatty
acids as a mol% and using novel clustering analysis approaches, no individual or
specific fatty acid species in the NEFA fraction predicted lower beta-cell
function, insulin sensitivity, or dysglycemia status. Taken together, these
results suggest that it is the absolute size of the circulating NEFA fraction,
irrespective of any specific composition of fatty acid species, that is
responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.

Biologically, within a free-living population chronic elevation of NEFA may be 
mediating it's association through TG. In normal metabolism, NEFA enters the 
liver and assists in the production of TG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating it's association
through TG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are some possible explanations for these findings. Previous studies done *in
vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.

There are two other possible explanations for our null findings for the insulin
sensitivity measures. First, there may be differences in physiology between
fasting and postprandial NEFA kinetics. For instance, some experimental studies
using clamp protocols found that fasting NEFA was a weak predictor of insulin
sensitivity compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue following postprandial TG lipolysis via lipoprotein lipase may result in
NEFA spillover into the blood and a subsequent increase in circulating NEFA
[@Almandoz2013a], which may be more metabolically active given postprandial 
activity. Second, the null findings may be due to the high risk population
examined in PROMISE. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat mode advanced
stage in the pathogenesis of diabetes.

Few studies have examined the composition of NEFA on metabolic functioning. One
recent, well-analyzed study used a variety of advanced fatty acid measurement
and statistical techniques to explore the multivariate relationship between the
NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a].
Specifically, the authors identified NEFA
16:1n-9, 20:1n-9, and 22:4n-6 to correlate with components of the MetS. 
Another similar study examining diabetes found that 16:0, 18:0, 18:1, 18:2, 18:3
may be useful biomarkers for identifying healthy compared to diabetic
individuals [@Liu2010a]. However, in both studies, major limitations were the
smaller sample sizes (approximately 100 subjects) and the cross-sectional
design.

@@

```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results)
```

@@

```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly)
```

@@

>>>>

<!--

TODO:
- Use pct for PLSDA

@@

Title: Total non-esterified fatty acids, but not individual fatty acids, predict
progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

Word count: 3569 / 4000

@@

# Abstract
249 / 250 words

Objective: Our aim was to determine the longitudinal associations of individual
non-esterified fatty acids (NEFA) on changes in insulin sensitivity and
beta-cell function over 6 years in a cohort of individuals who are at-risk for
diabetes.

@@

Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
6 years. Our results suggest that association with beta-cell function is due to
the absolute size of the serum NEFA fraction, rather than the specific fatty acid
composition.

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- [X] Interaction by sex or ethnicity
- [X] Interaction by time
- [X] Sensitivity analysis with TAG
- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
- [X] Replace TAG with BaseTAG in the QIC (or use both)
- [X] Fix time axis labels on LCMM plot
- Describe LCMM and PLS-DA better in plots

@@

sfig_plsda_groups <- sfig('plsda_groups', '')

@@

sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', '')

@@

sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA). Discriminatory ability is evident from the amount of separation between the two groups. In this case, PLS-DA was 82% accurate at correctly classifying groups.')
sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006). PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis. Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006). The percent explained variance of each component is shown in brackets on each axis. Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status. See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence 
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
exposure to high concentrations of NEFA can induce insulin resistance in 
insulin-sensitive tissues such as muscle and liver, and in addition can impair 
pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
Observational and clinical studies have reported concordant findings, showing
in particular that elevated total plasma NEFA associates with an increased risk
for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].

Much of the previous experimental work on the role of NEFA in type 2 diabetes
utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
(18:1n-9), or alternatively used specific oils such as soybean oil, which is
high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
exposure to characterize the impact of total NEFA. However, fatty acids comprise
multiple molecules with diverse physiological functions, and few 
studies have compared the effects of a broader spectrum of fatty acids. Notably,
one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].

Despite a sizable literature studying the role of total NEFA concentration in 
diabetes, there are important gaps in the research. The majority of previous 
studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
been epidemiological studies that only looked at total NEFA and not individual
fatty acids [@Miller2012a;@Johns2014a]. To date, there have
been no longitudinal studies examining the role of the composition of the serum NEFA
fraction on the pathogenesis of diabetes, which is critical to
understanding associations at a more granular level given the protracted natural
history of diabetes and the growing appreciation of the divergent effects of
individual fatty acids. Therefore, our research objective was to examine the 
association of serum NEFA composition on changes over time in insulin
sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
that higher palmitic acid and lower polyunsaturated fatty acids such as
eicosapentaenoic acid would associate with declining insulin sensitivity and
beta-cell function over 6 years.

@@

At each examination, an 8-12 hour fasting blood sample was draw from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the 
baseline visit, which had been frozen at -70°C for 4-6 years and had not been 
exposed to any freeze-thaw cycles. Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
of heptadecanoic acid (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch 
[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
and TAG) was isolated using thin layer chromatography; each fraction was
visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
ionization detector, which were injected in splitless mode. Fatty acid
concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a]. 
There were 22 fatty acids measured in the NEFA fraction. Given the
diverse biology, as well as the complexity of the analyses, findings for other lipid
fractions in this cohort are reported separately (see {{pl and ce paper}} for the
analysis of the phospholipid and cholesteryl ester fractions).

@@

The primary outcome variables for this analysis were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this analysis were
22 individual NEFA as mole percent (mol%) of the total fraction and as
concentration (nmol/mL). Pearson correlation coefficients were computed to
assess the relationships of individual NEFA with other continuous variables.

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design 
an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
is well suited to longitudinal cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled
(mean centered and standardized). The NEFA variables were classified as
*time-independent* (held constant) as they were measured only at the baseline
visit, while the outcome variables and covariates were set as *time-dependent*.
No imputation was conducted on missing values.

Covariate selection was based on previous literature, directed acyclic graph 
[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
we selected differed between insulin sensitivity and beta-cell function models. 
For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
function, the last few models (M5-M8) were comparable in performance, so we
report results from the simplest model (M8), adjusting for time, sex, ethnicity,
baseline age, WC, ALT,
and family history of diabetes. After scaling, log-transforming, and
exponentiating the GEE estimates, the GEE results are interpreted as an expected
percent difference in the outcome variable for every SD increase in the
predictor variable given the covariates are held constant. While we a priori
decided to test for and report on the time by predictor interaction results, we
also assessed the overall fit of models that included a time interaction, which
ultimately did not improve the overall fit of our models in terms of QIC
(M6.1 and M8.1 as seen
in `r cite_st('qic')`) and we therefore did not include them in the main effect
analysis. Given that TAG is a risk factor
for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
between NEFA and the outcomes. To determine the role of TAG in the
association between NEFA and the outcomes, TAG was included in the GEE model as a
sensitivity analysis. Lastly for the GEE models, we tested for an interaction
with sex or ethnicity by the predictor term for each outcome variable.

@@

Several associations were seen in the unadjusted GEE models 
(see `r cite_sf('gee')`), particularly for the concentration compared to the
mol% data. Full model adjustment (`r cite_f('gee')`) attenuated most of these
associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
had negative associations between
IGI/IR and ISSI-2. The magnitude of association for each of these variables was
fairly consistent for each beta-cell function measure. For every one SD decrease
in any of these three NEFA variables, there was an average predicted
`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
clinic visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
(data not shown).
There were no significant interaction effects
between time, sex, or ethnicity and the individual NEFA on any of the outcome
measures (all p>0.15). Raw values from the GEE models are shown in 
`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
`r lcmm.groups$Freq[3]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
response variables in the PLS-DA analysis. The PLS-DA results show a very poor
discriminatory ability of the NEFA composition in classifying participants into
the correct beta-cell function latent class. Only 
`r misclass.n.percent.bcf$NPct[2]` 
participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function, however, those individuals 
classified into the low beta-cell function group tended to have patterns of NEFA 
composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of
the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
or maintenance of dysglycemia status over the 6 years as the response variable showed similar
poor discriminatory ability, though better than the beta-cell function
trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants
were classified correctly based on dysglycemia status, examining the clustering
of the NEFA composition showed poor discrimination between groups (see 
`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).

@@

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
total NEFA had a negative association with insulin sensitivity and a
particularly strong negative association with beta-cell function [@Johns2014a].
Experimentally, several potential mechanisms have been elucidated for the role
of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
through endoplasmic reticulum stress, formation of ceramides, and generation of
nitric oxide, as well as impairment of proinsulin production and mitochrondrial
function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
The present analysis is the first study to our knowledge to examine the 
longitudinal association of a broad spectrum of individual NEFA on
beta-cell function in a large cohort. We found that there was a strong signal of
higher total NEFA, palmitic acid, oleic acid, and linoleic acid as
concentrations with lower beta-cell function. However, in modeling these fatty
acids as a mol% and using novel clustering analysis approaches, no individual or
specific fatty acids in the NEFA fraction strongly predicted lower beta-cell
function, insulin sensitivity, or dysglycemia status. Taken together, these
results suggest that it is the absolute size of the circulating NEFA fraction,
irrespective of any specific composition of fatty acids, that is
responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.

Biologically, within a free-living population chronic elevation of NEFA may be 
mediating it's association through TAG. In normal metabolism, NEFA enters the 
liver and assists in the production of TAG that is to be processed into very-low
density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
to hypertriglyceridemia, which is a known risk factor for diabetes. In the
sensitivity analysis adjusting for TAG, all associations with beta-cell function
were attenuated, suggesting that NEFA may in fact be mediating it's association
through TAG.

There is substantial experimental evidence highlighting the role of increased 
NEFA and the subsequent increase in insulin resistance via impairment of insulin 
signaling cascades, as reviewed in previously published articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
longitudinal analysis, we saw no association of any individual or total
NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
There are some possible explanations for these findings. Previous studies done *in
vivo* were mostly short term infusion protocols
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional or correlational
[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
these studies and the experimentally-induced
elevations in NEFA concentrations make it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR. However, we 
could not examine this potential hypothesis as NEFA was only collected at the
baseline visit.

There are two other possible explanations for our null findings for the insulin
sensitivity measures. First, there may be differences in physiology between
fasting and postprandial NEFA kinetics. For instance, some experimental studies
using clamp protocols found that fasting NEFA was a weak predictor of insulin
sensitivity compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
NEFA spillover into the blood and a subsequent increase in circulating NEFA
[@Almandoz2013a], which may be more metabolically active given postprandial 
activity. Second, the null findings may be due to the high risk population
examined in PROMISE. It may be that in this population, IR has become well
established and NEFA may not contribute to IR at this somewhat mode advanced
stage in the pathogenesis of diabetes.

Few studies have examined the composition of NEFA on metabolic functioning. One
recent, well-analyzed study used a variety of advanced fatty acid measurement
and statistical techniques to explore the multivariate relationship between the
NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a].
Specifically, the authors identified NEFA
16:1n-9, 20:1n-9, and 22:4n-6 to correlate with components of the MetS. 
Another similar study examining diabetes found that 16:0, 18:0, 18:1, 18:2, 18:3
may be useful biomarkers for identifying healthy compared to diabetic
individuals [@Liu2010a]. However, both studies were limited by
smaller sample sizes (approximately 100 subjects) and the cross-sectional
design.

@@

```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results, 'Beta-cell function')
```

@@

```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
```

@@

```{r supFigPlsExample, fig.cap=sfig_plsda_example}
plot_plsda_grouping(example_plsda_results(), 'Example groups')
```
